---
title: "Visualize incomplete and imputed data with `ggmice`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{mice}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4
)
```

# Overview

The package `ggmice` bundles a set of plotting functions for the imputation package `mice`. These plotting functions adhere to the 'grammar of graphics' style, popularized by the `ggplot2` package. With that, `ggmice` enhances imputation workflows and provides plotting objects that are easy to extend and manipulate. 

The main `mice` package contains several `lattice` style plotting functions, `ggmice` provides `ggplot2` versions of these functions and more. The core function in the `ggmice` package, the `ggmice()` function, can be applied to obtain `ggplot2` objects for both incomplete *and* imputed data. This vignette provides an overview of the main functions included in `ggmice`.

# Set-up

The `ggmice` package can be installed from GitHub as follows:

```{r install, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("amices/ggmice")
```

It is recommended to load the imputation package `mice` and the plotting package `ggplot2` into your workspace as well.

```{r setup}
library(ggmice)
dat <- mice::boys
imp <- mice::mice(dat, method = "pmm", printFlag = FALSE)
```

# Incomplete data

The `ggmice` package contains functions to explore incomplete data. 

## Missing data pattern

```{r pattern, echo=TRUE, eval=FALSE}

```

## Correlations between variables

The function `plot_corr()` can be used to investigate relations between variables, for the development of imputation models. Only one of the arguments (`dat`, the incomplete dataset) is required, all other arguments are optional.

```{r correlations}
plot_corr(dat)
plot_corr(dat, vrb = c("hgt", "wgt", "bmi"), label = TRUE, square = FALSE, diagonal = TRUE)
```

## Predictor matrix

The function `plot_pred()` displays `mice` predictor matrices. A predictor matrix is typically created using `mice::make.predictorMatrix()`, `mice::quickpred()`, or by using the default in `mice::mice()` and extracting the `predictorMatrix` from the resulting `mids` object. The `plot_pred()` function only requires the `pred` argument.

```{r predictormatrix}
pred <- mice::quickpred(dat)
plot_pred(pred)
plot_pred(pred, label = TRUE, square = FALSE)
```

## The `ggmice()` function

The `ggmice` function processes incomplete data in such a way that it can be displayed with `ggplot2`. The missing values are displayed on the axes.

```{r incomplete}
ggmice(dat, ggplot2::aes(age, bmi)) +
  ggplot2::geom_point()
ggmice(dat, ggplot2::aes(gen, bmi)) +
  ggplot2::geom_point()
```

# Imputed data

The `ggmice` package contains functions to evaluate observed and imputed data. 

## Algorithmic convergence

```{r convergence}
plot_chains(imp, "bmi")
```

## Box and whiskers plot

```{r bwplot}
bwplot(imp, "bmi")
```

## Stripplot

```{r stripplot}
stripplot(imp, "bmi")
```

## Densityplot

```{r densityplot}
densityplot(imp, "bmi")
```

## X-Y plot

```{r xyplot}
xyplot(imp, "age", "bmi")
```

## The `ggmice()` function

```{r imputed}
ggmice(imp, ggplot2::aes(log(wgt), hgt)) +
  ggplot2::geom_point()
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) + 
  ggplot2::geom_jitter() +
  ggplot2::geom_boxplot(size = 1, fill = "white", alpha = 0.5, outlier.shape = NA)
```

