---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{mice}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4
)
# TODO add to vignette:
# plotting conditional on missingness indicator
# adding jitter to categorical variables
# plotting a single imp
# plotting all variables
```

# Visualize incomplete and imputed data with `ggmice`

The `ggmice` package provides plotting functions for the evaluation of incomplete data, `mice` imputation models, and multiply imputed data sets (`mice::mids`). The functions in `ggmice` adhere to the 'grammar of graphics' philosophy, popularized by the `ggplot2` package. With that, `ggmice` enhances imputation workflows and provides plotting objects that are easily extended and manipulated by each individual 'imputer'. 

This vignette gives an overview of the core plotting functions in `ggmice`. Experienced `mice` users may already be familiar with the `lattice` style plotting functions in `mice`. These 'old friends' such as `mice::bwplot()` can be re-created with `ggmice`, see [this](https://amices.org/ggmice/articles/old_friends.html) vignette for advise.

<!-- Loading `ggmice` after `mice` masks these functions and provides suggestions for `ggplot2` equivalents these 'old friends'.  -->
<!-- The `mice` package contains several plotting functions that may guide 'imputers' in their workflow.  `lattice` style plotting functions, `ggmice` provides `ggplot2` versions of these functions and more. The core function in the `ggmice` package, the `ggmice()` function, can be applied to obtain `ggplot2` objects for both incomplete *and* imputed data. This vignette provides an overview of the main functions included in `ggmice`. -->

# Set-up

The `ggmice` package can be installed from GitHub as follows:

```{r install, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("amices/ggmice")
```

In this vignette we'll use `ggmice` in combination with the imputation package `mice` and the plotting package `ggplot2`. It is recommended to load `mice` and `ggplot2` into your workspace as well, but in your own workflow you could choose to call their functions directly instead (e.g., `mice::mice()` or `ggplot2::aes()`). In this vignette, we assume that all three packages are loaded, as well as an incomplete and imputed version of the `mice::boys` dataset.

```{r setup}
# load packages
library(mice)
library(ggplot2)
library(ggmice)

# load incomplete dataset 
dat <- boys

# generate imputations
imp <- mice(dat, method = "pmm", printFlag = FALSE)
```

# Incomplete data

The `ggmice` package contains functions to explore incomplete data. 

### Missing data pattern

The `plot_pattern()` function displays the missing data pattern in an incomplete dataset. The argument `data` (the incomplete dataset) is required, the argument `square` is optional and determines whether the missing data pattern has square or rectangular tiles, and the optional argument `rotate` changes the angle of the variable names 90 degrees if requested.

```{r pattern}
# create missing data pattern plot
plot_pattern(dat)

# specify optional arguments  
plot_pattern(dat, square = TRUE, rotate = TRUE)
```

### Influx and outflux

The `plot_flux()` function produces an influx-outflux plot. The influx of a variable quantifies how well its missing data connect to the observed data on other variables. The outflux of a variable quantifies how well its observed data connect to the missing data on other variables. In general, higher influx and outflux values are preferred when building imputation models. The plotting function requires an incomplete dataset (argument `data`), and takes optional arguments to adjust the legend and axis labels. 

```{r flux}
# create influx-outflux plot
plot_flux(dat)

# specify optional arguments  
plot_flux(
  dat, 
  label = FALSE, 
  caption = FALSE
)
```


### Correlations between variables

The function `plot_corr()` can be used to investigate relations between variables, for the development of imputation models. Only one of the arguments (`data`, the incomplete dataset) is required, all other arguments are optional.

```{r correlations}
# create correlation plot
plot_corr(dat)

# specify optional arguments  
plot_corr(
  dat,
  vrb = c("hgt", "wgt", "bmi"),
  label = TRUE,
  square = FALSE,
  diagonal = TRUE
)
```

### Predictor matrix

The function `plot_pred()` displays `mice` predictor matrices. A predictor matrix is typically created using `mice::make.predictorMatrix()`, `mice::quickpred()`, or by using the default in `mice::mice()` and extracting the `predictorMatrix` from the resulting `mids` object. The `plot_pred()` function requires a predictor matrix (the `data` argument), but other arguments can be provided too.

```{r predictormatrix}
# create predictor matrix
pred <- quickpred(dat)

# create predictor matrix plot
plot_pred(pred)

# specify optional arguments  
plot_pred(
  pred, 
  label = TRUE, 
  square = FALSE
)
```

### The `ggmice()` function

The `ggmice` function processes incomplete data in such a way that it can be displayed with `ggplot2`. The missing values are displayed on the axes (i.e., a missing value for the x-variable is plotted on top of the y-axis, and vice versa). Note that, in contrast to the `ggplot()` function, `ggmice()` *requires* an aesthetic mapping (argument `mapping`).

```{r incomplete}
# create scatter plot with continuous variables
ggmice(dat, aes(age, bmi)) +
  geom_point()

# create scatter plot with a categorical variable
ggmice(dat, aes(gen, bmi)) +
  geom_point()
```

# Imputed data

The `ggmice` package contains two functions to evaluate observed and imputed data. 

### Algorithmic convergence

The function `plot_trace()` plots the trace lines of the MICE algorithm for convergence evaluation. The only required argument is `data` (to supply a `mice::mids` object). The optional argument `vrb` defaults to `"all"`, which would display traceplots for all variables. 

```{r convergence}
# create traceplot for one variable
plot_trace(imp, "bmi")
```

### The `ggmice()` function

The `ggmice` function is versatile. It produces a `ggplot` object that can be extended to mimic every type of plot for observed and imputed data in `mice`, see see [this](https://amices.org/ggmice/articles/old_friends.html) vignette for advise. Below are some examples of plots produced with `ggmice()`. Note that, in contrast to the `ggplot()` function, `ggmice()` *requires* an aesthetic mapping (argument `mapping`).

```{r imputed}
# create scatter plot with continuous variables
ggmice(imp, aes(age, bmi)) +
  geom_point()

# create scatter plot with a categorical variable
ggmice(imp, aes(gen, bmi)) +
  geom_point()

# create scatter plot with a transformed variable
ggmice(imp, aes(log(wgt), hgt)) +
  geom_point()

# create stripplot with boxplot overlay
ggmice(imp, aes(x = .imp, y = bmi)) + 
  geom_jitter(height = 0) +
  geom_boxplot(fill = "white", alpha = 0.75, outlier.shape = NA) +
  labs(x = "Imputation number")
```

___

```{r session, class.source = 'fold-hide'}
# this vignette was generated with R session
sessionInfo()
```

