---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{mice}
  %\VignetteDepends{ggplot2}
---

<style>
p.caption {
  font-size: 0.7em;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

<!-- pkgdown::build_article("ggmice") -->

<!-- # Introduction -->

The R package `ggmice` is designed to support multiple imputation workflows with visualizations. 
`ggmice` provides 'grammar of graphics' (`ggplot2`) functionality for exploring incomplete data, building `mice` imputation models, and evaluating multiply imputed datasets (`mids` objects). 
 
## Minimal example

Set-up the environment in R with packages for imputation and visualization.
```{r, warning = FALSE, message = FALSE}
library(mice) 
library(ggplot2)
library(ggmice)
```

Visualize the `boys` data set from the `mice` package with `ggmice`. 
This is an incomplete dataset ($n = 748$) with cross-sectional data on $9$ growth-related variables (e.g., age in years and height in cm). 
```{r, fig.cap = "Scatterplot of height (in cm) by age (in years) in the incomplete `boys` dataset. Missing values are visualized on the axes lines, showing that some cases have observed data for age but missing height values."}
ggmice(
	boys,
	aes(age, hgt)) + 
	geom_point()
```

The blue points are observed, missing values are plotted on top of the axes in red.

Impute the missing values in the `boys` data set with `mice`.
```{r, warning = FALSE, message = FALSE}
imp <- mice(
	boys,
	m = 1,
	print = FALSE)
```

Visualize the imputed `boys` data set with `ggmice`.
```{r, fig.cap = "Scatterplot of height (in cm) by age (in years) in the imputed `boys` dataset. Missing values in the height variable are filled in with `mice` and visualized with `ggmice`."}
ggmice(
	imp,
	aes(age, hgt)) + 
	geom_point()
```

This figure shows the observed data in blue and the imputed data in red.
The missing data have been imputed along the range of the observed data, with some uncertainty due to the missingness.
These visualizations offer direct graphical comparisons between incomplete and imputed data, and may be used to evaluate imputation model fit.

## Outline

In this vignette, you will learn how to create and interpret `ggmice` visualizations for each phase of your imputation workflow. 

The vignette shows how to use `ggmice` to:

- explore incomplete data,
- build and inspect imputation models, and
- evaluate imputation models and imputed data.

But first, we need some background information about the `ggmice` package and its core function, `ggmice()`.

## The `ggmice` package

The `ggmice` package is designed to unify the visualization of incomplete and imputed data. 
It is an extension package to the popular imputation package `mice`.

`mice` has become the de facto standard in R to handle the ubiquitous problem of incomplete data. 
With `mice`, missing data points are ‘imputed’ (i.e. filled in) to obtain several completed data sets. 
Filling in the missing data multiple times allows for a valid representation of the uncertainty due to missingness.
`ggmice` offers tools to build and evaluate `mice` imputation models.

The functions in `ggmice` adhere to `ggplot2`'s 'grammar of graphics' philosophy. 
The resulting plots are standard `ggplot` objects, so you can add layers, labels, themes, facets, and transformations as usual.
These editable `ggplot` objects are easily adapted into publication-quality graphics.

<!-- `ggmice` supports imputation workflows and provides plotting objects that can be manipulated by each individual 'imputer'.  -->
<!-- Use `ggmice` to inspect missing data, develop imputation models, evaluate algorithmic convergence, or compare observed versus imputed data. -->
<!-- The resulting `ggplot` objects may be easily manipulated or extended.  -->

<!-- The vignette gives an overview of the different plotting functions in `ggmice`.  -->

## The `ggmice()` function

The core function in the `ggmice` package is `ggmice()`. 
`ggmice()` is a `ggplot2::ggplot()` wrapper function, which plots either missing or imputed values. 

- When the `data` argument is supplied with an incomplete dataset (a `data.frame` object), `ggmice()` visualizes observed versus *missing* data.
- When the `data` argument is supplied with multiply imputed datasets (a `mids` object), `ggmice()` visualizes observed versus *imputed* data.

By convention, observed data are plotted in blue and missing or imputed data in red.

The `ggmice()` function mimics how the `ggplot2` function `ggplot()` works.
Both take a `data` argument and a `mapping` argument, and will return an object of class `ggplot`. 

Using `ggmice()` looks equivalent to a `ggplot()` call:

```{r gg, eval=FALSE}
ggplot(boys, aes(x = age))
ggmice(boys, aes(x = age))
```

The main difference between the two functions is that `ggmice()` includes some pre-processing steps for incomplete and imputed data. 
The functions can be used interchangably, except for two key details:

1. The object supplied to the `data` argument in `ggmice()` may be either an incomplete dataset of class `data.frame`, or an imputation object of class `mice::mids`. 

Depending on whether a `data.frame` or `mids` object is provided, the resulting visualization will either contain observed and *missing* data, or observed and *imputed* data. 
By convention, observed data is plotted in blue and missing or imputed data is plotted in red. 

2. The `mapping` argument in `ggmice()` cannot be empty. 

Because of the internal processing in `ggmice()`, the `mapping` argument is *required* for each `ggmice()` call. 
This is in contrast to the aesthetic mapping in `ggplot()`, which may also be provided in subsequent plotting layers. 
An `x` or `y` mapping (or both) has to be supplied for `ggmice()` to function. 
This aesthetic mapping can be provided with the `ggplot2` function `aes()` (or equivalents). 
Other mapping may be provided too, except for `colour`, which is already used to display observed versus missing or imputed data.

After creating a `ggplot` object, any desired plotting layers may be added (e.g., with the family of `ggplot2::geom_*` functions), or adjusted (e.g., with the `ggplot2::labs()` function). 
This makes `ggmice()` a versatile plotting function for incomplete and/or imputed data. 

<!-- Loading `ggmice` after `mice` masks these functions and provides suggestions for `ggplot2` equivalents these 'old friends'.  -->
<!-- The `mice` package contains several plotting functions that may guide 'imputers' in their workflow.  `lattice` style plotting functions, `ggmice` provides `ggplot2` versions of these functions and more. The core function in the `ggmice` package, the `ggmice()` function, can be applied to obtain `ggplot2` objects for both incomplete *and* imputed data. This vignette provides an overview of the main functions included in `ggmice`. -->

### Incomplete data

If the object supplied to the `data` argument in the `ggmice()` function is a `data.frame`, the visualization will contain observed data in blue and missing data in red. 
Because missing data points are, by definition, unobserved, their values cannot be plotted directly. 
For categorical variables we can display the missing values as their own category.
For continuous variables, there is no straightforward way to show where the missingness occurs.

Visualizations of the missingness that we *can* create are via plots of variable pairs. 
Any missing values in one variable can be displayed on the axis of the other. 
This provides a visual cue that the missing data is distinct from the observed values, but still displays the observed value of the other variable. 

In the minimal example, the variable `age` was completely observed, while there were some missing entries for the height variable `hgt`. 
We create a scatter plot of these two variables with: 

```{r inc-con}
ggmice(boys, aes(age, hgt)) +
  geom_point()
```

The `age` of cases with missing `hgt` are plotted on the horizontal axis. 
This is in contrast to a regular `ggplot()` call with the same arguments, which would leave out all cases with missing `hgt`. 
So, with `ggmice()` we lose less information, and may even gain valuable insight into the missingness in the data.

### Imputed data

If the `data` argument in `ggmice()` is provided a `mice::mids` object, the resulting plot will contain observed data in blue and imputed data in red. 
<!-- There are many possible visualizations for imputed data, four of which are explicitly defined in the `mice` package.  -->
<!-- Each of these can be re-created with the `ggmice()` function (see the  [Old friends](https://amices.org/ggmice/articles/old_friends.html) vignette).  -->
<!-- But `ggmice()` can do even more.  -->

In the minimal example, we created the same scatter plot with the imputed data as with the incomplete data:

```{r imp-con, warning = FALSE}
ggmice(imp, aes(age, hgt)) +
  geom_point()
```

Aside from this minimal example, `ggmice` offers many more functionalities and functions.
For example, the package contains plotting functions to explore incomplete data and evaluate convergence of the imputation algorithm. 

Experienced `mice` users may already be familiar with the `lattice` style plotting functions in `mice` for visualizing imputed data. 
These 'old friends' such as `mice::stripplot()` can be re-created with the `ggmice()` function, see the [Old friends](https://amices.org/ggmice/articles/old_friends.html) vignette for advice.

In this vignette, the `ggmice` functions are presented in the order of a typical imputation workflow, where the missingness is first investigated, then imputation models are built based on relations between variables, and finally the imputations are inspected visually.

<br>

# Set-up

<!-- ## Packages -->

You can install the latest `ggmice` release from [CRAN](https://CRAN.R-project.org/package=ggmice) with:

``` r
install.packages("ggmice")
```

The development version of the `ggmice` package can be installed from GitHub with: 

``` r
# install.packages("devtools")
devtools::install_github("amices/ggmice")
```

After installing `ggmice`, you can load the package into your `R` workspace. 
It is highly recommended to load the `mice` and `ggplot2` packages as well. 
This vignette assumes that all three packages are loaded:

```{r setup, warning = FALSE, message = FALSE}
library(mice)
library(ggplot2)
library(ggmice)
```

<!-- ## Data -->

We will use the `penguins` data for illustrations, available from the base R `datasets`. 
The `penguins` dataset contains incomplete observational data on penguin sightings ($n = 344$).
Each row represents a single penguin, and the nine variables describe its characteristics (e.g., species) and measurements (e.g., body weight in grams). 

In this vignette, we will treat the penguins data as our working example. 
The missingness is empirical, and the data are of mixed type (continuous and categorical variables).
With that, we have the necessary packages (`mice`, `ggplot2`, and `ggmice`) and an incomplete dataset (`penguins`) to start the full imputation workflow.

<!-- We load the incomplete data from the `mice` package with: -->
<!-- ```{r data} -->
<!-- dat <- boys -->
<!-- ``` -->

<!-- For the purpose of this vignette, we impute all incomplete variables $m = 3$ times with predictive mean matching as imputation method.  -->
<!-- Imputations are generated with: -->

<!-- ```{r imp, results = "hide"} -->
<!-- imp <- mice(dat, m = 3, method = "pmm") -->
<!-- ``` -->

<!-- We now have the necessary packages, an incomplete dataset (`dat`), and a `mice::mids` object (`imp`) loaded in our workspace. -->

We will first use `ggmice` to explore marginal and joint distributions of incomplete variables. 
Next, we will construct and inspect imputation models based on relations between the variables. 
Finally, we will visualize imputation algorithm and the resulting imputations to assess whether the imputation models yields plausible values and adequately reflect the uncertainty due to missingness.

# Exploring incomplete data

  <!-- - `plot_pattern()` -->
  <!-- - `plot_miss()` -->
  <!-- - `ggmice()` -->
  
In a `mice` workflow, the missing data should first be inspected before determining an imputation strategy.
<!-- The R package ggmice aids data analysts in exploring the missing parts of their data. -->
The missing data pattern shows where in the incomplete data the missing values occur.
And multivariate graphs may highlight the severity of the missing data problem with respect to the substantive scientific estimate(s).

## `plot_miss()`

The `plot_miss()` function facilitates the exploration of where the missingness occurs.
The result is the graphical equivalent to the missingness matrix `is.na(penguins)`.

```{r}
plot_miss(penguins)
```

The plot can be ordered by the missingness proportion. 

```{r}
plot_miss(penguins, ordered = TRUE)
```

Other optional function arguments are `square` (for square tiles), `grid` (for grid lines between cells), `rotate` (to display the column names at a 90 degree angle), and `vrb` (to plot only a subset of variables).

## `plot_pattern()`

The `plot_pattern()` function displays the missing data pattern in an incomplete dataset. 
The argument `data` is required, which specifies the incomplete dataset. 

```{r pattern}
# create missing data pattern plot
plot_pattern(penguins)
```

The argument `square` is optional, and determines whether the missing data pattern has square or rectangular tiles.
The optional argument `rotate` changes the angle of the variable names by 90 degrees if requested. 
Other optional arguments are `cluster` and `npat`.
```{r pattern2}
# specify optional arguments
plot_pattern(
  penguins,
  square = TRUE,
  rotate = TRUE,
  npat = 2,
  cluster = "island"
)
```

<!-- ## `plot_miss()` -->

## `ggmice()`

The `ggmice()` function can be used to explore the distributions of incomplete variables, associations with other variables, and with missingness indicators.

Plot the distribution of an incomplete variable with:
```{r}
ggmice(penguins, aes(x = sex)) +
  geom_bar(fill = " white")
```

We can create scatterplots of the incomplete data. 
For example, plot the bivariate distribution of two continuous incomplete variables with:
```{r}
ggmice(penguins, aes(x = bill_len, y = bill_dep)) +
  geom_point()
```

We can plot the incomplete continuous variable `bill_dep` against the incomplete categorical variable `sex` with:

```{r}
ggmice(penguins, aes(x = sex, y = bill_dep)) +
  geom_point()
```

In this graph, the missing values are plotted on the axes lines. 

- Cases with observed `bill_len` and missing `sex` are plotted on the vertical axis. 
- There are no cases with observed `sex` and missing `bill_len`, but otherwise these would be plotted on the horizontal axis. 
- Cases with missing `bill_len` as well as missing `sex` are plotted on the intersection of the two axes, since no data is available. 

The 'grammar of graphics' makes it easy to adjust the plots programmatically. 
We could convert the plotted values of the variable `bill_len` from centimeters to inches with:

```{r inc-trans, warning = FALSE}
ggmice(penguins, aes(x = sex, y = bill_dep / 2.54)) +
  geom_point() +
  labs(y = "Bill depth (inches)")
```

If we would be interested in the sex differences in bill length between the penguin species, we can just add facets based on a clustering variable with:

```{r inc-clus}
ggmice(penguins, aes(x = sex, y = bill_dep)) +
  geom_point() +
  facet_wrap(~ species, labeller = label_both)
```



# Building imputation models

  <!-- - `plot_corr()` -->
  <!-- - `plot_flux()` -->
  <!-- - `plot_pred()` -->
  <!-- - `ggmice()` -->
  
<!-- - Model building: how do we know which functional form to choose? -->
<!--   - rely on defaults, tested in simulation studies  -->
<!--   - assess the distribution of the incomplete variable (e.g., visual inspection) -->
<!--   - ...? (maybe ad transformations?) -->
<!-- - Model building: how do we know which imputation model predictors to select? -->
<!--   - association of incomplete variable with other variables in the data ["include an auxiliary variable if its correlation with the main variable is ≥0.5 (in absolute value) [25]" @nguyen2021] -->
<!--   - association of missingness indicator with other variables in the data  -->
<!--   - external input (e.g., branching patterns, or expert knowledge to correct for MNAR) -->
<!--   - Take the analysis model (if available) as starting point for imputation workflow, for congeneality -->
<!--   - ...? -->

Imputation with mice requires an imputation model for each incomplete variable.
The defaults in `mice` are to set an imputation method based on the column type, and to use all other variables as imputation model predictors.
For example, a numeric data column would get the semi-parametric imputation method 'predictive mean matching' (`pmm`).

```{r}
pred <- make.predictorMatrix(penguins)
meth <- make.method(penguins)
plot_pred(pred, method = meth, square = FALSE)
```

## `plot_corr()`

The function `plot_corr()` can be used to investigate relations between variables for the development of imputation models. 
Only one of the arguments (`data`; the incomplete dataset) is required, all other arguments are optional.

```{r correlations}
# create correlation plot
plot_corr(penguins)

# specify optional arguments
plot_corr(
  penguins,
  vrb = c("bill_len", "bill_dep", "sex"),
  label = TRUE,
  square = FALSE,
  diagonal = TRUE,
  rotate = TRUE
)
```

## `ggmice()`

When `ggmice()` is applied to incomplete data with faceting based on a missingness indicator, associations between observed and missing data become apparent. 
This may help explore the missingness mechanisms in the incomplete data. 
The distribution of the continuous variable `bill_dep` and categorical variable `species` are visualized faceted by the missingness indicator for `sex` with:

```{r}
# continuous variable
ggmice(penguins, aes(bill_dep)) +
  geom_density() +
  facet_wrap(~ factor(is.na(sex) == 0, labels = c("sex observed", "sex missing")))
```

We can see differences in the distribution of `bill_dep` based on the missingness indicator of `sex`, so `bill_dep` is informative for imputing `sex`.

```{r}
# categorical variable
ggmice(penguins, aes(species)) +
  geom_bar(fill = "white") +
  facet_wrap(~ factor(is.na(sex) == 0, labels = c("sex observed", "sex missing")))
```

In this case, we can see differences in the distribution of `species` based on the missingness indicator of `sex`, so `species` is informative for imputing `sex`.

## `plot_flux()`

The `plot_flux()` function produces an influx-outflux plot.
The influx of a variable quantifies how well its missing data connect to the observed data on other variables. 
The outflux of a variable quantifies how well its observed data connect to the missing data on other variables. 
In general, higher influx and outflux values are preferred when building imputation models. 
The plotting function requires an incomplete dataset (argument `data`), and takes optional arguments to adjust the legend and axis labels.

```{r flux}
# create influx-outflux plot
plot_flux(penguins)

# specify optional arguments
plot_flux(
  penguins,
  label = FALSE,
  caption = FALSE
)
```



## `plot_pred()`

The function `plot_pred()` displays `mice` predictor matrices. 
In a `mice` workflow, the predictor matrix determines which imputation model predictors will be used when imputing the incomplete variables.
There are three ways to create a predictor matrix programmatically with `mice`.
The first and most explicit way is using `mice::make.predictorMatrix()`. 
The second way is implicitly, by using the default in `mice::mice()` and extracting the `predictorMatrix` from the resulting `mids` object. 
The third way is via `mice::quickpred()`, which selects the most relevant imputation model predictors based on associations in the data.
The `plot_pred()` function requires a predictor matrix (the `data` argument), but other arguments can be provided too.

```{r predictormatrix}
# create predictor matrix
pred <- quickpred(penguins)

# create predictor matrix plot
plot_pred(pred)

# specify optional arguments
plot_pred(
  pred,
  label = FALSE,
  square = FALSE,
  rotate = TRUE,
  meth = meth
)
```

# Imputing

Run the imputation algorithm.

```{r}
imp <- mice(penguins, pred = pred, meth = meth, print = FALSE, seed = 123)
```

# Evaluating imputations

  <!-- - `plot_trace()` -->
  <!-- - `ggmice()` -->

<!-- - Model evaluation: how do we assess imputation model misfit? -->
<!--   - fit of the model to the observed data -->
<!--   - non-convergence in the algorithm  -->
<!--   - mismatch between observed and imputed data distributions (e.g., visual inspection) [@buur18] -->
<!--   - ...? -->

## `plot_trace()`

The function `plot_trace()` plots the trace lines of the MICE algorithm for convergence evaluation. 
The only required argument is `data` (to supply a `mice::mids` object). 
The optional argument `vrb` defaults to `"all"`, which would display traceplots for all variables. 

```{r convergence}
# create traceplot for all variables
plot_trace(imp)
```

Add iterations to the imputation algorithm.

```{r}
imp <- mice.mids(imp, maxit = 5, print = FALSE)
```

```{r}
plot_trace(imp)
```

## `ggmice()`

We can create the same scatter plots as the ones on the incomplete data, but now on the imputed data:

```{r imp-same, warning = FALSE}
ggmice(imp, aes(x = sex)) +
  geom_bar(fill = " white")
ggmice(imp, aes(x = bill_len, y = bill_dep)) +
  geom_point()
ggmice(imp, aes(x = sex, y = bill_dep)) +
  geom_point()
ggmice(imp, aes(x = sex, y = bill_dep / 2.54)) +
  geom_point() +
  labs(y = "Bill depth (inches)")
ggmice(imp, aes(x = sex, y = bill_dep)) +
  geom_point() +
  facet_wrap(~ species, labeller = label_both)
```

These figures show the observed data points once in blue, plus five imputed values in red for each missing entry.



It is also possible to use the imputation number as mapping variable in the plot. 
For example, we can create a stripplot of observed and imputed data with the imputation number `.imp` on the horizontal axis:

```{r imp-strip}
ggmice(imp, aes(x = .imp, y = bill_len)) +
  geom_jitter(height = 0, width = 0.25) +
  labs(x = "Imputation number")
```

A major advantage of `ggmice()` over the equivalent function `mice::stripplot()` is that `ggmice` allows us to add subsequent plotting layers, such as a boxplot overlay:

```{r imp-box}
ggmice(imp, aes(x = .imp, y = bill_len)) +
  geom_jitter(height = 0, width = 0.25) +
  geom_boxplot(width = 0.5, linewidth = 1, alpha = 0.75, outlier.shape = NA) +
  labs(x = "Imputation number")
```

You may want to create a plot visualizing the imputations of multiple variables as one object. This can be done by combining `ggmice` with the functional programming package `purrr` and visualization package `patchwork`. For example, we could obtain boxplots of different imputed variables as one object using:

```{r facet}
purrr::map(c("bill_len", "bill_dep", "body_mass"), ~ {
  ggmice(imp, aes(x = .imp, y = .data[[.x]])) +
    geom_boxplot() +
    labs(x = "Imputation number")
}) %>%
  patchwork::wrap_plots()
```

For further inspection, we can also split the plots by imputation number, which allows us to evaluate the bivariate distribution within each imputation.
```{r}
ggmice(imp, aes(x = sex, y = bill_dep)) +
  geom_point() +
  facet_wrap(~.imp)
```


<!-- To re-create any `mice` plot with `ggmice`, see the [Old friends](https://amices.org/ggmice/articles/old_friends.html) vignette.  -->



---

#

This is the end of the vignette. This document was generated using:

```{r session, class.source = 'fold-hide'}
sessionInfo()
```

## Acknowledgements

The `ggmice` package is developed with guidance and feedback from Gerko Vink, Stef van Buuren, and others. This project has received funding from the European Union’s Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.
