---
title: "old_friends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{old_friends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggmice)
imp <- mice::mice(mice::boys, method = "pmm", printFlag = FALSE)
```

How to re-create the plotting functions from `mice` with `ggmice`, in alphabetical order.

# `mice::bwplot()`

Box-and-whisker plot of observed and imputed data.

```{r bwplot}
# original plot
mice::bwplot(imp, bmi ~ .imp)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
      ggplot2::geom_boxplot() +
      ggplot2::labs(x = "Imputation number")
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
  ggplot2::stat_boxplot(geom = 'errorbar', linetype = "dashed") +
  ggplot2::geom_boxplot(outlier.colour = "grey", outlier.shape = 1) +
  ggplot2::labs(x = "Imputation number")
```

# `mice::densityplot()`

Density plot of observed and imputed data.

```{r densityplot}
# original plot
mice::densityplot(imp, ~bmi)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = bmi, group = .imp)) +
      ggplot2::geom_density() 
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(x = bmi, group = .imp, size = .where)) +
      ggplot2::geom_density() +
      ggplot2::scale_size_manual(values = c("observed" = 1, "imputed" = 0.5), guide = "none")
```

# `mice::flux()`

[To be added]

# `mice::md.pattern()`

Missing data pattern plot.

```{r md.pattern}
# original plot
mice::md.pattern(imp$data)
# ggmice equivalent
plot_pattern(imp$data)
```

# `mice::plot.mids()`

Plot the trace lines of the MICE algorithm.

```{r plot.mids}
# original plot
plot(imp, bmi ~ .it | .ms)
# ggmice equivalent
plot_trace(imp, "bmi")
```

# `mice::stripplot()`

Stripplot of observed and imputed data.

```{r stripplot}
# original plot
mice::stripplot(imp, bmi ~ .imp)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
      ggplot2::geom_jitter(width = 0.25) +
      ggplot2::labs(x = "Imputation number")
# extended reproduction with ggmice

```

# `mice::xyplot()`

Scatterplot of observed and imputed data.

```{r}
# original plot
mice::xyplot(imp, bmi ~ age)
# ggmice equivalent
ggmice(imp, ggplot2::aes(age, bmi)) +
    ggplot2::geom_point()
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(age, bmi)) +
    ggplot2::geom_point(shape = 1)
```

