---
title: "Old friends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{old_friends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Create the `ggmice` equivalent of `mice` plots

How to re-create the output of the plotting functions from `mice` with `ggmice`. In alphabetical order of the `mice` functions.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4
)
```

First load the `ggmice` package, some incomplete data and a `mice::mids` object into your workspace. 

```{r setup}
# load packages
library(ggmice)
# load incomplete dataset 
dat <- mice::boys
# generate imputations
imp <- mice::mice(dat, method = "pmm", printFlag = FALSE)
```


# `bwplot`

Box-and-whisker plot of observed and imputed data.

```{r bwplot}
# original plot
mice::bwplot(imp, bmi ~ .imp)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
      ggplot2::geom_boxplot() +
      ggplot2::labs(x = "Imputation number")
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
  ggplot2::stat_boxplot(geom = 'errorbar', linetype = "dashed") +
  ggplot2::geom_boxplot(outlier.colour = "grey", outlier.shape = 1) +
  ggplot2::labs(x = "Imputation number") +
  ggplot2::theme(legend.position = "none")
```

# `densityplot`

Density plot of observed and imputed data.

```{r densityplot}
# original plot
mice::densityplot(imp, ~bmi)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = bmi, group = .imp)) +
      ggplot2::geom_density() 
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(x = bmi, group = .imp, size = .where)) +
  ggplot2::geom_density() +
  ggplot2::scale_size_manual(values = c("observed" = 1, "imputed" = 0.5),
                             guide = "none") +
  ggplot2::theme(legend.position = "none")
```

# `fluxplot`

Influx and outflux plot of multivariate missing data patterns.

```{r flux}
# original plot
mice::fluxplot(dat)
# ggmice equivalent
plot_flux(dat)
```


# `md.pattern`

Missing data pattern plot.

```{r md.pattern}
# original plot
md <- mice::md.pattern(dat)
# ggmice equivalent
plot_pattern(dat)
# extended reproduction with ggmice
plot_pattern(dat, square = TRUE) +
  ggplot2::theme(legend.position = "none",
                 axis.title = ggplot2::element_blank(),
                 axis.title.x.top = ggplot2::element_blank(),
                 axis.title.y.right = ggplot2::element_blank())
```

# `plot.mids`

Plot the trace lines of the MICE algorithm.

```{r plot.mids}
# original plot
plot(imp, bmi ~ .it | .ms)
# ggmice equivalent
plot_trace(imp, "bmi")
```

# `stripplot`

Stripplot of observed and imputed data.

```{r stripplot}
# original plot
mice::stripplot(imp, bmi ~ .imp)
# ggmice equivalent
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
  ggplot2::geom_jitter(width = 0.25) +
  ggplot2::labs(x = "Imputation number")
# extended reproduction with ggmice (not recommended)
ggmice(imp, ggplot2::aes(x = .imp, y = bmi)) +
  ggplot2::geom_jitter(
    shape = 1,
    width = 0.1,
    na.rm = TRUE,
    data = data.frame(
      bmi = dat$bmi,
      .imp = factor(rep(1:imp$m, each = nrow(dat))),
      .where = "observed"
    )
  ) +
  ggplot2::geom_jitter(shape = 1, width = 0.1) +
  ggplot2::labs(x = "Imputation number") +
  ggplot2::theme(legend.position = "none")
```

# `xyplot`

Scatterplot of observed and imputed data.

```{r}
# original plot
mice::xyplot(imp, bmi ~ age)
# ggmice equivalent
ggmice(imp, ggplot2::aes(age, bmi)) +
  ggplot2::geom_point()
# extended reproduction with ggmice
ggmice(imp, ggplot2::aes(age, bmi)) +
  ggplot2::geom_point(size = 2, shape = 1) +
  ggplot2::theme(legend.position = "none")
```

___

```{r session, class.source = 'fold-hide'}
# this vignette was generated with R session
sessionInfo()
```
